\section{Threads}
\label{sec:threads}

We can identify a particular kind of a lens-like structure which may arise in a ledger specification,
which we call a \emph{thread}, and denote $T \in \Th$. More specifically,

\begin{definition}
  A \emph{thread} $T \in \Th$ in a ledger specification $L$ is a tuple of
  \begin{itemize}
    \item[(i)] an underlying type $T$,

    \item[(ii)] a projection function
    \[ \fun{proj}_T ~\in~\type{State}_L \to T^? \]

    That is surjective, with a right inverse $\fun{inj}_T$ :

    \[ \fun{proj}_T~\circ~\fun{inj}_T ~=~\fun{id}_{T^?} \]

    \item[(iii)] an update function $\fun{updateT}_T$ is such that
    for all $tx,~s$,

    \[ \fun{updateT}_T~tx~s~(\fun{proj}_T~s) = \Err~ \Rightarrow~ \ups~tx~s~=~ \Err \]

    If $\ups~tx~s~\neq~ \Err$,

    \[ \fun{updateT}_T~tx~s~(\fun{proj}_T~s) = \fun{proj}_T~(\ups~tx~s)\]
  \end{itemize}
\end{definition}

Note that the projection function is not from a valid state, but rather from an
arbitrary state. This function can be extended to a map between dependent types :

\[ \fun{depProj}_T ~\in~\type{State}_L \to T_{val} \]

where

\[ T_{val}~\leteq~\{ ~(s,~t)~ \mid~\fun{proj}_T~s~=~t~\wedge~s~\in~\type{ValSt}_L~ \} \]

Note also that the projection function may return $\Nt$, which implies that all threads
in a ledger are non-persistent. While this is not true for most realistic ledgers
(eg. the UTxO set thread always exists on the ledger, even if it is empty), a ledger specification may
restrict the collection of valid states to only those where persistent threads
are never projected to $\Nt$.

The function $\fun{updateT}$ which applies a transaction-state pair to a $t \in T$
is defined in terms of the $\ups$ function in Figure \ref{fig:threads}.

Recall the one-object category than the one-object category $\type{SLC}_L$.

We construct the category as follows :



 where each $v \in T$ is an image of some $s \in S$.
  and the hom-set of $T \to T$ is given by pairs of type $(\type{ValSt}_{L},~\seqof{(\type{Tx}_{L})})$.

  The update function used to apply a list of transactions $(s,~txs)$ is given by
  folding $\fun{updateT}$ over $txs$ at $s$.
  That is, maps in this category are specified by both the transactions being applied to the ledger,
  as well as the ledger state itself.

  \todopv{Can we do pointed pairs here (i.e. $(t, s)$) instead as objects, rather than just the thread?}


Note here that the implication is one direction --- if the $\fun{updateT'}_T$
does not produce an error, the $\ups$ function can still produce an error applied
to the same transaction and state. That is, intuitively, a thread update function
only cares that the transaction is attempting to
update the thread data badly, and not about what it does to the rest of the ledger.

Since a $\Err$ in the output of the $\fun{updateT'}_T$ function implies in an error
in the $\ups$ function applied to the same state and transaction,
$\fun{updateT'}_T$ is said to \emph{emit constraints} on the ledger update
function.



If we consider individual stateful smart contracts as threads, we again see that
the update of the total UTxO set (or the account-based ledger's state) require
that some general rules must be satisfied, while each smart contract will
have its own constraints that it may place on the transaction performing its
state update.

\begin{itemize}
  \item[(i)] A \emph{temporary} thread $T$ is one that may not appear in some valid ledger
  states of a ledger specification $\type{Spec}~T = L$, i.e.

  \[ \exists~~s~\in~\type{ValSt}_L,~\fun{proj}_T~s~=~\Nt \]

  A permanent thread is one that is not temporary.

  \item[(ii)] A \emph{stateful} thread $T$ is one where the projection function,
  for some pair of ledger states, gives two different non-$\Nt$ outputs.
  \[ \exists~~s,~s'~\in~\type{ValSt}_L,~\Nt~\neq~\fun{proj}_T~s~\neq~\fun{proj}_T~s'~\neq~\Nt \]

  \item[(iii)] We say that the domain of a thread $T$ with the update function
  $\fun{updateT}_T$ \emph{constrains} the
  update of the ledger state when for some $tx$,

  \[ \fun{dom}_T~(tx,~s)~\neq~\fun{id}_T \]

  \item[(iv)] Threads $T$ and $U$ in a ledger specification $L~=~\type{Spec}_T~=~\type{Spec}_U$
  are said to be \emph{independent} when any change by a transaction $tx$ to the state of both threads can
  be instead implemented by two lists of transactions $txs1, txs2$, where

  \begin{itemize}
    \item applying the
    list $txs1$ to the ledger state updates one of the threads to have the same state as after applying
    $tx$ to the ledger while keeping the state of the other thread constant, and
    \item applying the list $txs2$ to the resulting ledger state updates the state of the
    other thread to be the same as after applying $tx$.
  \end{itemize}

  \[ \forall~tx~s~\in~\type{ValSt}_L,~~\fun{validTx}_L~tx~s~\Rightarrow~~\exists~~txs1~txs2, \]
  \[ \fun{proj}_{T}~(\fun{updateErr}_{L}~tx~s)~=~\fun{proj}_{T}~(\fun{computeState}_L~s~(txs1++txs2)) \]
  \[ \wedge~ \fun{proj}_{U}~(\fun{updateErr}_L~tx~s)~=~\fun{proj}_{U}~(\fun{computeState}_L~s~(txs1++txs2)) \]
  \[ \wedge \]
  \[ \fun{proj}_{T}~(\fun{computeState}_L~s~txs1)~=~\fun{proj}_{T}~(\fun{updateErr}_L~tx~s)  \]
  \[ \wedge~ \fun{proj}_{U}~(\fun{computeState}_L~s~txs1)~=~\fun{proj}_{U}~s \]

  \item[(iv)] Threads $T$ and $U$ in a ledger specification $L~=~\type{Spec}~T~=~\type{Spec}~U$
  are said to be \emph{separateable by message passing} when there exist threads $T'$ and $U'$
  such that
  \[ \forall~tx~s~\in~\type{ValSt}_L,~~\fun{validTx}_L~tx~s \]
  ......
\end{itemize}

\todopv{when can two dependent threads be made separable?}

\subsection{Deterministic threads}

\todopv{Does this work?}

See Figures \ref{fig:order-thread}, \ref{fig:update-thread}. UD defined in \ref{sec:update}.

\begin{figure*}[htb]
  \emph{Order-determinism for a thread $T$ in ledger $L$}
  %
  \begin{equation*}
    \begin{array}{l@{~~}l@{~~}ll}
      \forall~(s~\in~\type{ValSt}_L)~(txs~\in~\seqof{\type{Tx}_L}),~\\
      & ~~~~txs'~\in~\fun{Permutation}~txs,~\\
      & ~~~~\fun{validUpdate}_L~s~txs~\neq~\Err~\neq~\fun{validUpdate}_L~s~txs' \\
      & ~~~~\Rightarrow~~\fun{proj}_{T}~(\fun{validUpdate}_L~s~txs)~=~\fun{proj}_{T}~(\fun{validUpdate}_L~s~txs')
    \end{array}
  \end{equation*}
  \caption{Order-determinism for a single thread}
  \label{fig:order-thread}
\end{figure*}

\begin{figure*}[htb]
  \emph{Update-determinism for a thread $T$ in ledger $L$}
  %
  \begin{equation*}
    \begin{array}{l@{~~}l@{~~}ll}
      % \forall~s~tx~tx',~\\
      % & ~~~~ (\Err~\neq~(s~\var{tx})~=~(s~\var{tx'})~\neq~\Err~\Rightarrow~\\
      % & ~~~~\forall~s',~(s'~\var{tx})~\neq~\Err~\neq~(s'~\var{tx'}) \\
      % & ~~~~~~~~\Rightarrow~\fun{proj}_{T}~(s'~\var{tx})~=~\fun{proj}_{T}~(s'~\var{tx'}))
    \end{array}
  \end{equation*}
  \caption{Consistent transaction application (UD) for a single thread $T$}
  \label{fig:update-thread}
\end{figure*}
