\section{Threads}
\label{sec:threads}

So far, we have not identified any structure in the underlying state types of
ledgers. To study specific parts of a ledger, we introduce a concept we call
a \emph{thread}, and denote $T \in \Th$. More specifically,

\begin{definition}
  A \emph{thread} $T \in \Th$ in a ledger specification $L$ is a tuple of
  \begin{itemize}
    \item[(i)] an underlying type $T$,

    \item[(ii)] a function
    \[ \fun{proj}_T ~\in~\type{State} \to T \]

    That such that it has a right inverse $\fun{inj}_T$ :

    \[ \fun{proj}_T~\circ~\fun{inj}_T ~=~\fun{id}_T \]

    \item[(iii)] an update function $\fun{updateT}_T$ is such that
    for all $tx,~s$,

    \[ \fun{updateT}_T~tx~s~(\fun{proj}_T~s) = \Err~ \Rightarrow~ \ups~tx~s~=~ \Err \]

    And if $\ups~tx~s~\neq~ \Err$,

    \[ \fun{updateT}_T~tx~s~(\fun{proj}_T~s) = \fun{proj}_T~(\ups~tx~s)\]
  \end{itemize}
\end{definition}

Note that the projection function is not from a valid state, but rather from an
arbitrary state.

\textbf{Relation between updating the thread and ledger states. }

In (iii), the implication is one direction --- if the $\fun{updateT}_T$
does not produce an error for a given $s, tx$, the $\ups$ function can still produce an error applied
to the same transaction and state. That is, intuitively, a thread update function
only cares that the transaction is attempting to
update the thread data badly, and not about what it does to the rest of the ledger.

Since a $\Err$ in the output of the $\fun{updateT}_T$ function implies in an error
in the $\ups$ function applied to the same state and transaction,
$\fun{updateT}_T$ is said to \emph{emit constraints} on the ledger update
function.

Given any type $T$ and the $\fun{proj}_T$ function, the function $\fun{updateT}$, which
applies a transaction-state pair to a $t \in T$,
can be defined in terms of the $\ups$ function by specifying that, unconditionally,

\[ \fun{updateT}_T~tx~s~(\fun{proj}_T~s) = \fun{proj}_T~(\ups~tx~s)\]

\textbf{One-object categories reperesenting threads. }

Recall the one-object category than the one-object category $\type{SLC}_L$, with
lists of transactions as the functions. We can construct a related one-object category $\mathcal{T}$ for
any thread $T$ as follows :

\begin{itemize}
  \item the object $T_{val}$ is the image of $\ValSt$ under the $\fun{proj}_T$ function,
  \item the hom-set of $T_{val} \to T_{val}$ is given by the set of pairs
  $(s,~txs) : (\type{ValSt}_{L},~\seqof{(\type{Tx}_{L})})$

  \item the update function used to apply a list of transactions $(s,~txs)$ is given by
  folding $\fun{updateT}$ over $txs$ at $s$
\end{itemize}

For $\fun{proj}_T = \fun{id}_{\State}$, the $\mathcal{T}$ is the same as $\type{SLC}_L$.

\todopv{Can we do pointed pairs here (i.e. $(t, s)$) instead as objects, rather than just the thread?}

\begin{itemize}
  \item[(i)] A \emph{non-persistent} thread $T~\cong~T'^?$ is one that may take on an $\Nt$ value in some valid ledger
  states of a ledger specification $L$, i.e.

  \[ \exists~~s~\in~\type{ValSt},~\fun{proj}_T~s~=~\Nt \]

  A thread is called \emph{persistent} otherwise.

  \item[(ii)] A \emph{single-state} thread $T$ is one where the projection function,
  for some pair of ledger states, outputs exactly one non-$\Nt$ value.
  \[ \forall~~s,~s'~\in~\type{ValSt}_L,~\fun{proj}_T~s~\neq~\Nt~\neq~\fun{proj}_T~s' \]

  \[ \Rightarrow~\fun{proj}_T~s~=~\fun{proj}_T~s' \]

\end{itemize}


\textbf{Deterministic threads}

\todopv{Does this work?}

Order-determinism for a single thread is defined in Figure \ref{fig:order-thread}.

\begin{figure*}[htb]
  \emph{Order-determinism for a thread $T$ in ledger $L$}
  %
  \begin{equation*}
    \begin{array}{l@{~~}l@{~~}ll}
      \forall~(s~\in~\type{ValSt})~(txs~\in~\seqof{\type{Tx}}),~\\
      & ~~~~txs'~\in~\fun{Permutation}~txs,~\\
      & ~~~~(s~txs)~\neq~\Err~\neq~(s~txs') \\
      & ~~~~\Rightarrow~~\fun{proj}_{T}~(s~txs)~=~\fun{proj}_{T}~(s~txs')
    \end{array}
  \end{equation*}
  \caption{Order-determinism for a single thread}
  \label{fig:order-thread}
\end{figure*}

Since a ledger is itself a thread, the following lemma follows immediately :

\begin{lemma}
  A ledger is OD if and only if all of its threads are.
\end{lemma}

A (probably stronger) notion of ledger determinism expressed via threads, see Figure \ref{fig:det-thread}.
This notion of determinism for threads implies thread order-determinism.


\begin{figure*}[htb]
  \emph{Determinism for a thread $T$ in ledger $L$}
  %
  \begin{equation*}
    \begin{array}{l@{~~}l@{~~}ll}
      \forall~(s~s'~\in~\type{ValSt})~(tx~\in~\type{Tx}),~\\
      & ~~~~(s~tx)~\neq~\Err~\neq~(s'~tx), \\
      & ~~~~\fun{proj}_{T}~s~=~\fun{proj}_{T}~s' \\
      & ~~~~\Rightarrow~~\fun{proj}_{T}~(s~tx)~=~\fun{proj}_{T}~(s'~tx)
    \end{array}
  \end{equation*}
  \caption{Determinism for a single thread}
  \label{fig:det-thread}
\end{figure*}

\subsection{Thread examples.}

\textbf{Binary-valued ledgers $L$ and $K$.}

Consider two ledgers, $L$ and $K$, such that

\[ \State_L = \Tx_L = \{ (a,~b)~\mid~a,~ b~ \in~ \{ 0,~1\}~\}  =  \State_K = \Tx_K \]

and $\fun{initState}_L = \fun{initState}_K = (0,~0)$, with update functions as follows :

\begin{itemize}
  \item[($L$)] $((a,b) ~(t_1,t_2))~ =~ (a~\fun{XOR}~t_1,~t_2)$

  \item[($K$)] $((a,b) ~(t_1,t_2))~ = \begin{cases}
    (a~\fun{XOR}~t_1,~t_2) & \text{if } b = t_2 \\
    \Err & \text{otherwise }
  \end{cases} $
\end{itemize}

The ledger $L$ is not order deterministic, since applying the transaction list $[ (0,1) ; (0,0) ]$
produces a different state than $[ (0,0) ; (0,1) ]$ :

\[ ((0,0) ~(0,1)~(0,0)) = (0, 0) \neq (0, 1) = ((0,0) ~(0,1)~(0,0)) \]

Each term of the tuple is updated using only data from the current state of that term, and
does not inspect the other term. We can express this in terms of threads, where $T \leteq \{ 0,~1 \}$,
and $\fun{proj}_T \leteq \pi_0$,

\[ \fun{updateT}~(a,b)~a'~(t_1,~t_2)~=~\begin{cases}
  a~\fun{XOR}~t_1 & \text{if } a = a' \\
  \Err & \text{otherwise }
\end{cases} \]

The first coordinate is updated using the $\fun{XOR}$ function, which is commutative.
The second coordinate also forms a thread, with $\fun{proj}_{T'} = \pi_1$, but it
is not deterministic.

\todopv{we want to say that the change in the first coordinate is always +~t1~ mod~2,
whereas the change in the second coordinate depends on the current state (i.e. either "flip the bit" or
"dont flip the bit", depending on whether b and t2 match), even though
the output itself does not depend on the current state}

The ledger $K$ also has threads formed by the first and second projection function,
however, both are deterministic. The $\pi_0$ thread is deterministic as in $L$, and
the $\fun{updateT}$ function for the projection $\pi_1$
either makes no changes, or fails -- instead of flipping the bit when $b \neq t_2$ (as in $L$) .

\textbf{List-of-transactions ledger $\type{SLC}_{list}$.}

Recall the ledger $\type{SLC}_{list}$ for a fixed transaction type $\Tx$, where the
state is a list of processed transactions. Note that ledger itself is not OD, since
reordering transactions produces distinct lists.

Every other ledger $T$ with the same transaction type $\Tx$ and initial state $i$ is a (possibly non-persistent)
thread in $\type{SLC}_{list}$, where $\fun{validUpdate}_T~i$ is the $\fun{proj}_T : [\Tx] \mapsto \State^?$
function, with $\Nt$ representing the $\Err$ state.

\subsection{Characterization of ledger determinism via threads.}

\todopv{I would like to find any additional preconditions P1..Pn such that it is possible to say that :
order-determinism for a single thread + P1...Pn implies thread determinism in the stronger sense --
hopefully, no additional preconditions P1..Pn are necessary}
